
# Reproducible Research - Assignment 1
## By Long Huynh
###### 19/12/2015
___

The purpose of this assignment is to demonstrate my ability to analyse data and present my findings using R Markdown and Knitr. I will be using a  [dataset](https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip) that was taken from an anonymous individual between October 2012 to November 2012, containing 3 variables:

* **Steps**: Number of steps taken in a 5-minute interval.
* **Date**: The date of which the measurement was taken in YYYY-MM-DD format.
* **Interval**: Identifier for the the 5-minute interval in which measurement was taken.

This document presents both the R Code and the results from my findings.

## Settings

One of the requirements for this assignment is to show all the code and the outputs. Therefore I set the entire code chunks in the document as `echo = TRUE`.

```{r}
library(knitr)
opts_chunk$set(echo = TRUE)

```

## Loading libraries

To support my analysis I initially loaded 3 libraries that made data importing (`readr`), wrangling (`dplyr`) and visualization (`ggplot2`) easier. All 3 packages were created by Hadley Wickham.

```{r echo}
## Loading libraries
library(dplyr)
library(readr)
library(ggplot2)
```

## Importing and cleaning the dataset

Before I began to answer the questions I used reproducible code to import the [dataset](https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip) online. I downloaded and unzipped the data (if the dataset did not sit in the work directory) and used the `readr` package to load the data into a table called `tblActivity`. 

```{r}
# Load tables up ----------------------------------------------------------

nmUrl <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip"

if(!file.exists("repdata_data_activity.zip")){
    download.file(nmUrl, "repdata_data_activity.zip")
}
if(!file.exists("activity.csv")){
    unzip("repdata_data_activity.zip")
}

tblActivity <- read_csv("activity.csv") %>% tbl_df()
```

## What is mean total number of steps taken per day?

*For this part of the assignment, you can ignore the missing values in the dataset.*

1. *Calculate the total number of steps taken per day*

I first removed all missing values from the table, grouped the [dataset](https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip) against the `date` variable and assigned the table to `tblStepsByDay`. The table highlights the total number of steps taken per day.

```{r}

tblStepsByDay <- tblActivity %>%
    filter(!is.na(steps)) %>%     #ignore missing values
    group_by(date) %>% 
    summarise(steps = sum(steps)) %>% #Calculate the total number of steps taken per day
    print()

```

2. *If you do not understand the difference between a histogram and a barplot, research the difference between them. Make a histogram of the total number of steps taken each day*

I used the `ggplot2` package to create a histogram of the total number of steps that were taken per day. I additionally placed two vertical lines indicating the locations of the mean (solid red line) and median lines (dashed purple line) on the histogram. As you'll see on the histogram both the mean and median lay very closely to one another which represents the midpoint on the dataset.

```{r}

chtStepsTakenNaRemoved <- tblStepsByDay %>% 
    ggplot(aes(steps)) +
    geom_histogram(binwidth = 1000, 
                   fill="light blue", 
                   colour="black") + # 
    xlab("Number of steps") + # add labels to axis
    ylab("Number of days") +
    ggtitle("Total number of steps taken per day (NA removed)") + # add main title
    theme_bw() +
    theme(plot.title = element_text(size = 16), # resize text
          axis.title.x = element_text(size = 12),
          axis.title.y = element_text(size = 12)) +
    geom_vline(xintercept = mean(tblStepsByDay$steps), colour="red", lwd = 0.5) + #Indicate the mean point
    geom_vline(xintercept = median(tblStepsByDay$steps), linetype = "longdash", 
                                   colour="purple", lwd = 0.5) + #Indicate median point
    guides(fill = guide_legend(reverse=TRUE))
chtStepsTakenNaRemoved


```

3. *Calculate and report the mean and median of the total number of steps taken per day*

I used the R base functions to identify the midpoints values from the dataset.

```{r}
meanStepsPerDay <- mean(tblStepsByDay$steps) %>% print()
medianStepsPerDay <- median(tblStepsByDay$steps) %>% print()
```

* **Mean Steps** : `r as.integer(meanStepsPerDay)`
* **Median Steps** : `r medianStepsPerDay`

## What is the average daily activity pattern?

1. *Make a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis).*

Before I made the time series plot I first needed to create an aggregated table for the average number of steps taken across all days. Unfortunately the original table time intervals was not in a time format. Therefore I used the `stringr` package to convert the `interval` variable to time.

```{r}

library(stringr)

tblAverageStepsTaken <- tblActivity %>% 
    mutate(hours = ifelse(str_count(interval)<=2,"00", str_pad(str_sub(interval, 1, 
                                                       str_count(interval)-2), 2, pad = "0")),
           minutes = str_pad(str_sub(interval, -2, -1), 2, pad = "0"),
           interval = paste(hours, ":", minutes, sep="")) %>% 
    group_by(interval) %>% 
    summarise(steps = mean(steps, na.rm = TRUE)) %>% 
    print()

```

I used the `ggplot2` package to visualize the data into a time series for a 24 hour period. I loaded the `scales` library to show time on the x axis.

```{r}
library(scales)

tblAverageStepsTaken %>%
    mutate(interval = as.POSIXct(interval, "%H:%M",tz = "GMT")) %>% 
    ggplot(aes(interval, steps)) +
    geom_line(colour = "red") +
    scale_x_datetime("5 Minute Intervals", labels = date_format("%H:%M")) +
    ylab("Steps per 5 min intervals") +
    ggtitle("Time series plot of the average number of steps against all days") +
    theme_bw()

```

 
2. *Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?*

```{r}

maxNumStepsInterval <- tblAverageStepsTaken %>% 
    filter(steps == max(tblAverageStepsTaken$steps)) %>% 
    print()

```

On average the `r maxNumStepsInterval$interval` interval had the highest number of steps, with `r as.integer(maxNumStepsInterval$steps)` steps.

## Imputing missing values

*Note that there are a number of days/intervals where there are missing values (coded as NA). The presence of missing days may introduce bias into some calculations or summaries of the data.*

1. *Calculate and report the total number of missing values in the dataset (i.e. the total number of rows with NAs)*
```{r}

totalNA <- tblActivity %>% 
    filter(is.na(steps)) %>% 
    nrow() %>% 
    print()

```

There are `r totalNA` missing values within the data set.

2. *Devise a strategy for filling in all of the missing values in the dataset. The strategy does not need to be sophisticated. For example, you could use the mean/median for that day, or the mean for that 5-minute interval, etc.*

For missing values I imputed the median number of steps against the 5 minute interval using the `ifelse` and `median` functions. This is highlighted in the code along with my response to the next question.

3. *Create a new dataset that is equal to the original dataset but with the missing data filled in.*

```{r}

stepsPerDayMedianImputed <- tblActivity %>% 
    group_by(interval) %>% 
    mutate(steps = ifelse(is.na(steps), median(steps, na.rm = TRUE), steps)) %>%
    group_by(date) %>% 
    summarise(steps = sum(steps)) %>% 
    print()

```

4. *Make a histogram of the total number of steps taken each day and Calculate and report the mean and median total number of steps taken per day. Do these values differ from the estimates from the first part of the assignment? What is the impact of imputing missing data on the estimates of the total daily number of steps?*

```{r}

chtStepsTakenMedianImputed <- stepsPerDayMedianImputed %>% 
    ggplot(aes(steps)) +
    geom_histogram(bin = 1000, colour = "black", fill = "light blue") +
    theme_bw() +
    ggtitle("Total number of steps taken per day (Median steps imputed)") +
    geom_vline(aes(xintercept = mean(stepsPerDayMedianImputed$steps)), colour="red", lwd = 0.5) + #Mean
    geom_vline(aes(xintercept = median(stepsPerDayMedianImputed$steps)), linetype = "longdash", colour="purple", lwd = 0.5) + #Median
    guides(fill = guide_legend(reverse=TRUE)) +
    xlab("Number of steps") +
    ylab("Number of days") +
    theme(plot.title = element_text(size = 16),
          axis.title.x = element_text(size = 12),
          axis.title.y = element_text(size = 12))
chtStepsTakenMedianImputed

```

I used `ggplot` to create a histogram similar to the one created in question 1. I also aligned two vertical lines indicating the mean (red solid line) and median (purple dashed line). When comparing the two plots you'll notice that majority of the imputed median days are shown against a lower interval (1000 to 2000 steps) which pulls the mean steps down. Please see the 2 histograms together below:

```{r}

library(gridExtra)
grid.arrange(chtStepsTakenNaRemoved, chtStepsTakenMedianImputed)

```

As you can see when comparing the two histograms together the mean value has fallen when imputing median values. There is now a wider variance between the mean and the median when imputing the median value.

```{r}
meanStepsPerDayImputed <- mean(stepsPerDayMedianImputed$steps)
medianStepsPerDayImputed <- median(stepsPerDayMedianImputed$steps)
```


Midpoint | NA Removed | Median Imputed | Difference
-------- | ---------- | -------------- | -----------
Mean | `r as.integer(meanStepsPerDay)` | `r as.integer(meanStepsPerDayImputed)` | `r as.integer(meanStepsPerDay) - as.integer(meanStepsPerDayImputed)`
Median | `r medianStepsPerDay` | `r medianStepsPerDayImputed` | `r medianStepsPerDay - medianStepsPerDayImputed`


## Are there differences in activity patterns between weekdays and weekends?

*For this part the `weekdays` function may be of some help here. Use the data set with the filled-in missing values for this part.*

1. *Create a new factor variable in the dataset with two levels -- "weekday" and "weekend" indicating whether a given date is a weekday or weekend day.*

I created a new factor variable called "Weekday" using the `mutate` function. If the day of week is either a Saturday or Sunday the `Weekday` variable would show "Weekend" otherwise it would show "Weekday". I then converted the new variable into a factor using the `as.factor` function.

```{r}
WeekdayvsWeekend_steps <- tblActivity %>% 
    mutate(hours = ifelse(str_count(interval)<=2,"00", str_pad(str_sub(interval, 1, 
                                                       str_count(interval)-2), 2, pad = "0")),
           minutes = str_pad(str_sub(interval, -2, -1), 2, pad = "0"),
           interval = paste(hours, ":", minutes, sep="")) %>% 
    group_by(interval) %>% 
    mutate(steps = ifelse(is.na(steps), median(steps, na.rm = TRUE), steps)) %>% 
    mutate(Day = weekdays(date), 
           Weekday = ifelse(Day %in% c("Saturday", "Sunday"), "Weekend", "Weekday"),
           Weekday = as.factor(Weekday)) %>% 
    select(steps, Weekday, interval) %>% 
    group_by(interval, Weekday) %>% 
    summarise(steps = mean(steps)) %>% 
    ungroup() %>% 
    print()
```

2. *Make a panel plot containing a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis).*

I used `ggplot` to create a time series plot with number of steps on the x axis and the mean number of steps along the y axis. I then faceted the chart across the `Weekday` factor.

```{r}
WeekdayvsWeekend_steps %>% 
    mutate(interval = as.POSIXct(interval, "%H:%M",tz = "GMT")) %>%  # Convert to time
    ggplot(aes(interval, steps)) +
    geom_line(aes(colour = Weekday)) +
    scale_x_datetime("5 Minute Intervals", labels = date_format("%H:%M")) +
    facet_grid(Weekday~.) + # facet over weekday
    theme_bw() +
    theme(legend.position = "none") +
    ggtitle("Mean number of steps taken. Weekdays vs Weekends") +
    xlab("Interval")+
    ylab("Mean number of steps") +
    theme(plot.title = element_text(size = 16),
          axis.title.x = element_text(size = 12),
          axis.title.y = element_text(size = 12))

```

The shapes of two plots are different. During the weekday the numbers steps tends to peak during the morning and evening (more so in the morning). Whereas on the weekend the number steps are active throughout the day. I suspect the subject is travelling to work in the morning and returning home in evening on weekdays. As the weekends does not follow the same pattern I believe the subject does not travel to work during Weekends.

_____



